An Approach to Off-line Handwriting Recognition through Segmentation Techniques and Deep Neural Networks

### 2017-12-18 Report/paper.tex
fix indentation

\begin{algorithm}[tb]
    \caption{Contour combination algorithm}
   \label{alg:overlap}
\begin{algorithmic}
    \STATE {\bfseries Input:}  $contour1$, $contour2$, $text$
    \STATE $xGap =$ XAxisDistance $contour1$, $contour2$
    \STATE $yOverlap =$ YAxisOverlap $contour1$, $contour2$
    \IF{ 
    \\ 
    \hspace{8px} $xGap < text$.averageWordWith 
    \\
    \hspace{8px} $* $ \hspace{2px} $(1 / (averageCharacterCount * 2))$ 
    \\
    {\bfseries and} 
    \\ 
    \hspace{8px} $yOverlap > min(contour1.height, contour2.height)$ 
    \\
    \hspace{8px} $* $ \hspace{2px} $  requiredYaxisOverlap $ 
    }
    \vspace{4px}
    \STATE combineContours $contour1$ $contour2$
    \ENDIF
\end{algorithmic}
The chosen paramters are:
\\
$averageCharacterCount = 8$ 
\\
$requiredYaxisOverlap = 1/2$
\end{algorithm}

@@

\begin{algorithm}[tb]
   \caption{Algorithm for line reconstruction}
   \label{alg:linereconstr}
\begin{algorithmic}
   \STATE {\bfseries Input:} $extractedWords$, $textlines$
   \REPEAT
        \STATE $newline = [ $wordWithHighestYPosition$ ]$
        \STATE $avgY= newline$.averageYPosition
        \REPEAT
            \FOR{$word$ {\bfseries in} $extractedWords$}
            \STATE $boundRect = word.contour.boundingRect$
            \IF{ $avgY > word.y$ {\bfseries and} $avgY < word.y + word.h$ }
            \STATE $line$.add $word$
            \ENDIF
        \ENDFOR
        \UNTIL{NotChangedLastIteration $line$}
        \STATE $textlines$.add $newline$
   \UNTIL{$words$.empty $= true$}
\end{algorithmic}
\end{algorithm}

@@

The next step in the process is to extract the character images from the found word images.
Again, there is a choice between several different approaches.
In this paper we focus solely on a word oversegmentation approach to character extraction, through skeletonization and using a vertical projection profile. The reasoning begind this can be found in the following papers \cite{CharSegm} and \cite{CharSegmOld}.

After first oversegmenting the word, we will recover the original character through postprocessing steps combining different word segments.

\subsection{Normalization}
The first action is again normalizing the word image.
We will resuze the image to a fixed height of $100px$.
This will result in more consistency in rule-based post processing steps.

\subsection{Binarization}
The word image is binarized.
This is again done with a combination of Gaussian blur and Otsu's method.
The binarization is needed for the next step, because we can only apply a skeletonization algorithm to a binary image.

\subsection{Skeletonization}
Now we will skeletonize our word image.
This will normalize the thickness of the strokes to a single pixel.
This allows us to execute our Verical projection profile and interpret the results easily.
The skeletonization algorithm used is the Zhang Suen thinning algorithm \cite{zsthinning}.

\subsection{Vertical projection histogram}
Using the skeletonized word image, we now create a vertical projection histogram.
This vertical projection histogram takes the sum of every collumn in the image, counting black pixels as a $1$ value, and white pixels as a $0$ value.

We now regard the collumns with a total pixel count of zero or one pixel as potential segmentation points.
We interpret these collumn to be the whitespace in between characters or the line connecting two characters to each other.
This will inevitably cause oversegmentation in our word image.
The next step will be to filter the potential segmentation points, and remove the incorrect ones through postprocessing.

@@

\subsection{Oversegmentation correction}
In this section we provide techniques to verify the correctness of segmentation points.
\cite{evalsplitpoints} discusses three approaches to evaluate segmentation point correctness: rule based techniques, cost function optimization based methods and machine learning based procedures.
In this paper we discuss machine learning based approaches precedented by a few rule based techniques.

\subsubsection{Rule based techniques}
The main oversegmentation correction is handled by our neural network.
However some simple preprocessing is added to remove obvious unwanted segmentation points.

We first remove removes the segmentation points found at the start and the end of the word image.
These segmentation points are an artifact of the used segmentation approach, as can be observed in Figure Figure \ref{fig:char_segm}.
These points can be removed without information loss, as the alphabet does not contain letters that entirely convert to a single pixel line after skeletonization.

Subsequently we group consequent segmentation points.
The occurence of these consequent segmentation points is another artifact from our segmentation approach, as can be seen in Figure \ref{fig:char_segm}.

From these groups of points, we take the center point as the new potential segmentation points.
However if one or more of the segmentation points in the group stems from empty collumns, the new segmentation point is chosen as the center point of the points representing the empty collumns in the word image.

Finally we aggregate groups of consequent potential segmentation points, separated by collumns containing less pixels than a certain threshold value $maxSumThresh$.
We have chosen a conservative value of $3px$ for this threshold.
This remedies the segmentation of slated lines inside characters, which can be an artifact of the skeletonization algirithm used.

A technique was tested using the feedback of the chracter recognition network to remove incorrect segmentation points.
However this turned out not to be ineffective, because the neural network tends to provide random probabilities with a high degree of certainty when images of multiple characters together were fed to the nerual network.
Since it is impossible to validate these responses from the neural network, this approach was abandoned and is not used in the final version.

More rule based filter could be added for more optimal processing of the potential segmentation points.
However the focus of our segmentation correction is the neural network used to remove incorrect segmenatation points.

@@

The text to word segmentation process can be improved upon and made more robust.
More advanced text segmentation approaches can be utilized to extract the text segments from documents containing images and other unwanted elements. \cite{textsegm}
Algorithms that provide better noise reduction can be added to handle text images with more noise,
and experiments with rotations of the text can provide us with a more robust system to tackle text images containing heavily slanted text lines.

The character segmentation process is insufficient as it is, and generally unable to segment words adequately so that the character recognition step can recognise the found segments.
More postprocessing can be used through rule based techniques and a more robust neural network to correct the found segmentation points. \cite{segmtechniques}

Finally, other approaches can also be considered.
State-of-the-art systems currently use segmentation-free approaches, this involves handwriting recognition without explicitly splitting words into characters. These methods are mostly based on hidden Markov models and neural networks.
This concept is explained and applied by \cite{presham}.
\section*{Acknowledgements}
We thank Sarah Vluymans for her comments on this paper.
\bibliography{paper}
\bibliographystyle{icml2016}

>>>>

\begin{algorithm}[tb]
    \caption{Contour combination algorithm}
    \label{alg:overlap}
\begin{algorithmic}
    \STATE {\bfseries Input:}  $contour1$, $contour2$, $text$
    \STATE $xGap =$ calculateXaxisDistance $contour1$, $contour2$
    \STATE $yOverlap =$ calculateYoverlap $contour1$, $contour2$
    \STATE $avgWidth =$ getAverageWordWith $text$
    \IF{ 
    $xGap < avgWidth $ \\ $* (1 / (averageCharacterCount * 2))$ \\  {\bfseries and} \\ $yOverlap > min(contour1.height, contour2.height)$ \\ $*  requiredYaxisOverlap $ 
    }
    \STATE combineContours $contour1$ $contour2$
    \ENDIF
\end{algorithmic}
The chosen paramters are:
\\
$averageCharacterCount = 8$ 
\\
$requiredYaxisOverlap = 1/2$
\end{algorithm}

@@

\begin{algorithm}[tb]
   \caption{Algorithm for line reconstruction}
   \label{alg:linereconstr}
\begin{algorithmic}
   \STATE {\bfseries Input:} $extractedWords$, $textlines$
   \REPEAT
        \STATE $highestWord = findHighestWord $
        \STATE $newline = [highestWord]$
        \STATE Initialize $noChange = true$.
        \STATE $avgY= newline$.findAverageYpos
        \REPEAT
            \FOR{$word$ {\bfseries in} $extractedWords$}
            \STATE $boundRect = word.contour.boundingRect$
            \IF{ $avgY > word.y$ {\bfseries and} $avgY < word.y + word.h$ }
            \STATE $line$.add $word$
            \STATE $noChange = false$
            \ENDIF
        \ENDFOR
        \UNTIL{$noChange = true$}
        \STATE $textlines$.add $newline$
   \UNTIL{$words$.empty $= true$}
\end{algorithmic}
\end{algorithm}

@@

The next step in the process is to extract the character images from the found word images.
Possible approaches and their advantages are discussed by \cite{CharSegm} and \cite{CharSegmOld}.
We choose the word oversegmentation approach for character extraction. Word oversegmentation is achieved through skeletonization and using a vertical projection profile which is discussed in consequently \ref{sec:skel} and \ref{sec:vertproj}. 
After oversegmenting the word, we find the actual splitpoints by removing unnecessary splitpoints as explained in \ref{sec:postprocesscharseg}

\subsection{Preprocessing}
We resize the image to a fixed height of 100 pixels. 
Next we binarize the image using a combination of Gaussian blur and Otsu's method. 
The binarization method is identical to \ref{sec:binartext} but included to allow segmentation of words which are not part of a text. Binarization is essential as it is necessary for skeletonization.

\subsection{Skeletonization}
\label{sec:skel}
Next we skeletonize the word. This normalizes the thickness of the strokes to a single pixel and
allows us to create a vertical projection profile. We use the skeletonization algorithm of \cite{zsthinning}.

\subsection{Vertical projection histogram}
\label{sec:vertproj}

Using the skeletonized word image, we create a vertical projection histogram.
This vertical projection histogram counts the black pixels of every column in the image.
We regard the columns with a total pixel count of zero or one as potential segmentation points.
We interpret these columns as whitespace in between characters or as a line connecting two characters to each other.
This will inevitably cause oversegmentation in our word image.
The next step filters incorrect segmentation points caused by oversegmentation.

@@

\subsection{Oversegmentation correction}
\label{sec:postprocesscharseg}

In this section we provide techniques to verify the correctness of segmentation points.
\cite{evalsplitpoints} discusses three approaches to evaluate segmentation point correctness: rule based techniques, cost function optimization based methods and machine learning based procedures.
In this paper we discuss machine learning based approaches precedented by a few rule based techniques.
An example of applying these correction techniques to an image is shown in \ref{fig:char_segm}.

\subsubsection{Rule based techniques}

First we remove segmentation points found at the start and the end of the word image.
These points can be removed without information loss, as the alphabet does not contain letters that entirely convert to a single pixel line after skeletonization.
Next consequent segmentation points are grouped. The center of these points is considered as segmentation point.
Unless if one or more of the segmentation points in the group stems from empty columns, the new segmentation point is chosen as the center of the points representing the empty columns in the word image.
Finally we aggregate groups of consequent potential segmentation points, separated by columns containing less pixels than a threshold value 3.
This remedies the segmentation of slanted lines inside characters.

A technique was tested using the output of the character recognition network to remove incorrect segmentation points.
This turned out to be ineffective because the neural network tends to provide high probabilities given images containing multiple characters. These probabilities are not meaningful and cannot be used as feedback for segmentation correctness.

@@

The text to word segmentation process can be improved upon and made more robust.
Advanced text segmentation approaches can be utilized to extract the text segments from documents containing images and other unwanted elements.
Algorithms that provide better noise reduction can be added to handle text images with more noise,
and experiments with rotations of the text can provide us with a more robust system to tackle text images containing heavily slanted text lines.

Finally, other approaches can also be considered.
State-of-the-art systems currently use segmentation-free approaches, this involves handwriting recognition without explicitly splitting words into characters. These methods are mostly based on hidden Markov models and neural networks. Due to the ambiguous nature of the word segmentation problem, a segmentation-free approach results in much better results.
This concept is explained and applied by \cite{presham}.
\section*{Acknowledgements}
We thank Sarah Vluymans for her comments on this paper.
\bibliography{paper}
\bibliographystyle{icml2016}
